Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-12-19 sections/combinatorial_semantics.tex
TODOs added

Unlike SMC equations the equations for semilattice enriched categories are not absorbed by e-hypergraph representation and hence to further build a correspondence between $\textsf{PROP}^{+}(\Sigma)$ and $\WellTypedMdaEcospans$ we add them as structural rewrite rules.
Recall that the category $\catname{EHyp_{\Sigma}}$ does not have all pushouts and hence we need to explicitly show that the pushouts that we need exist.

@@

>>>>

Unlike SMC equations the equations for semilattice enriched categories are not absorbed by e-hypergraph representation and hence to further build a correspondence between $\textsf{PROP}^{+}(\Sigma)$ and $\WellTypedMdaEcospans$ we will need to turn the latter into a semilattice-enriced category.
We already have a $+$ in $\WellTypedMdaEcospans$ so we need to require that it satisfies a particular set of equations, for example, that $h \otimes (f + g) = h \otimes f + h \otimes g$.
\update{How to formally state this? Because $h \otimes (f+g)$ is an SMC term from $\WellTypedMdaEcospans$ which is just an SMC so far.}
We will add such equations as DPOI rewrite rules and show that for each such rule the corresponding pushout complement and the pushout exist.

@@

\update{How do we state the correspondence between term equations and their e-hypergraph counterpart without a functor?}

### 2023-12-19 sections/soundness_and_completeness.tex
TODOs added

\[
[\cdot] : \textsf{PROP}^{+}(\Sigma) \to \WellTypedMdaEcospans
\]

by freeness, it suffices to define how the functor acts on generators.
As $\textsf{PROP}(\Sigma)$ and $\textsf{PROP}^{+}(\Sigma)$ essentially have the same set of generators we can reuse the functor $\llbracket \cdot \rrbracket: \textsf{PROP}(\Sigma) \to \MdaCospans$ from~\cite{Frobenius2} and then follow the inclusion $\mathcal{I}$ defined in~\ref{functor:I}.
Everything else follows by defining

\begin{enumerate}
    \item $[ f;g ] = [ f ] ; [ g ]$  for $f,g \in \Sigma$;
    \item $[ f \otimes g ] = [ f ] \otimes [ g ]$  for $f,g \in \Sigma$;
    \item $[ f + g ] = [ f ] + [ g ]$  for $f,g \in \Sigma$;
\end{enumerate}

this also makes the functor into a semilattice-enriched SMC one on the nose.
% \update{This essentially means that $[\cdot] = \llbracket \cdot \rrbracket ; \mathcal{I}$ plus three equations above.}

@@

\begin{proposition}
    Functor $\llbracket \cdot \rrbracket; \mathcal{I}$ is faithful.
    \begin{proof}
        Consider the commutative diagram below.

@@

\begin{theorem}[$\llbracket - \rrbracket$ is faithful]
\begin{proof}
Essentially we need to show that if $\llbracket f \rrbracket = \llbracket g \rrbracket$ then $f = g$ which is equivalent to if $f \not = g$ then $\llbracket f \rrbracket \not = \llbracket g \rrbracket$.
First, we note that we have a \question{weak (modulo commutativity)} normal form for any term in $PROP^{+}(\Sigma)$ which is achieved by applying distributivity for tensor and composition as well as unpacking equations, that is, every term is representable as in Figure~\ref{fig:normal_form_string} where $\phi_1,\;\ldots\;,\phi_n$ are pure SMC terms.
\update{It is not actaully a `normal' form because we can still apply SMC equations, so it should be normal modulo SMC and commutativity}.

@@

where all $A_1\; \ldots \;A_n$ do not contain any hierarchical edges.
We will write such forms as $\phi_1 + \ldots \phi_n$ and $A_1 + \ldots A_n$ respectively.
Then, $\llbracket f \rrbracket \not = \llbracket g \rrbracket$ leads to two options

% \begin{itemize}
%     \item $\llbracket f \rrbracket$ = $A_1 + \ldots A_n$ and $\llbracket g \rrbracket = B_1 + \ldots B_m$ where $n \not = m$. By functoriality and weak normal form in $\textsf{PROP}^{+}(\Sigma)$, $f = a_1 + \ldots + a_n$, $g = b_1 + \ldots b_m$ and $\llbracket f \rrbracket = \llbracket a_1 \rrbracket + \ldots + \llbracket a_n \rrbracket$, $\llbracket g \rrbracket = \llbracket b_1 \rrbracket + \ldots \llbracket b_m \rrbracket$. This essentially means that $f \not = g$ as it has different number of summands. 
%     \item $\llbracket f \rrbracket$ = $A_1 + \ldots A_n$ and $\llbracket g \rrbracket = B_1 + \ldots B_n$. Then there exists a permutation $\sigma$ and an index $i$ such that $ A_i  \not =  B_{\sigma(i)}$. By functoriality there exists $a_i$ such that $\llbracket a_i \rrbracket = A_i$ and $b_j$ such that $\llbracket b_j \rrbracket = B_{\sigma(i)}$. As $A_i$ does not have any hierarchical edges it effectively lives in $\mathcal{I}(\catname{MACsp_{D}(Hyp_{\Sigma})})$. By faithfulness of $\llbracket \cdot \rrbracket;\mathcal{I}$ it follows that $a_i \not = b_j$ and $f \not = g$. \update{Prove that $\llbracket \cdot \rrbracket; \mathcal{I}$ is faithful?}
% \end{itemize}

Now, assume that $f = \phi_1 + \ldots + \phi_n$ and $g = \psi_1 + \ldots + \psi_m$.
Then, by functoriality, $[f] = [A_1] + \ldots + [A_n]$ and $g = [B_1] + \ldots + [B_m]$. 
$f \not g$ gives us two options.

@@

\begin{theorem}

    Let $\mathcal{R}$ be a rewriting system on $\textsf{PROP}(\Sigma)$.
    $f \Rightarrow{}_{\mathcal{R}^{+}}$ iff $[f] \Rrightarrow{}_{[\mathcal{R}^{+}]} [g]$.
    \begin{proof}
        \update{It is tempting to reuse a similar result for $\llbracket \cdot \rrbracket$ because according to our definition $f \Rrightarrow_{\mathcal{R}^{+}}$ means that there is an SMC piece $f'$ inside some box that rewrites to $g'$.
            But we only have this result for $\llangle \cdot \rrangle$ for terms with bent wires which is a functor to $PROP + Frob$.
        }
    \end{proof}
\end{theorem}

>>>>

\[
[\cdot] : \textsf{PROP}^{+}(\Sigma) \to \sfrac{\WellTypedMdaEcospans}{\mathcal{S}}
\]
where $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ is $\WellTypedMdaEcospans$ quotiented by structural rewrite rules which makes this category into a semilattice-enrhiched one.
By freeness, it suffices to define how the functor acts on generators \update{Identities and generators from the signature?}.
% As $\textsf{PROP}(\Sigma)$ and $\textsf{PROP}^{+}(\Sigma)$ essentially have the same set of generators we can reuse the functor $\llbracket \cdot \rrbracket: \textsf{PROP}(\Sigma) \to \MdaCospans$ from~\cite{Frobenius2} and then follow the inclusion $\mathcal{I}$ defined in~\ref{functor:I}.
Everything else follows by defining

\begin{enumerate}
    \item $[ f;g ] = [ f ] ; [ g ]$  for $f,g \in \Sigma$;
    \item $[ f \otimes g ] = [ f ] \otimes [ g ]$  for $f,g \in \Sigma$;
    \item $[ f + g ] = [ f ] + [ g ]$  for $f,g \in \Sigma$;
    \item \update{Symmetry?}
\end{enumerate}

this also makes the functor into a symmetric monoidal semilattice-enriched SMC one on the nose.
% \update{This essentially means that $[\cdot] = \llbracket \cdot \rrbracket ; \mathcal{I}$ plus three equations above.}

@@

\begin{proposition}
    Functor $\llbracket \cdot \rrbracket; \mathcal{I}$ is faithful.
    \begin{proof}
        Consider the commutative diagram below \update{Is it obvious that the diagram commutes?}.

@@

\begin{theorem}[$\llbracket - \rrbracket$ is faithful]
\begin{proof}
Essentially we need to show that if $\llbracket f \rrbracket = \llbracket g \rrbracket$ then $f = g$ which is equivalent to if $f \not = g$ then $\llbracket f \rrbracket \not = \llbracket g \rrbracket$.
First, we note that we have a \question{weak (modulo commutativity)} normal form for any term \update{string diagram?} in $PROP^{+}(\Sigma)$ which is achieved by applying distributivity for tensor and composition as well as unpacking equations, that is, every term is representable as in Figure~\ref{fig:normal_form_string} where $\phi_1,\;\ldots\;,\phi_n$ are pure SMC terms.
\update{It is not actaully a `normal' form because we can still apply SMC equations, so it should be normal modulo SMC and commutativity}.

@@

where all $A_1\; \ldots \;A_n$ do not contain any hierarchical edges.
We will write such forms as $\phi_1 + \ldots \phi_n$ and $A_1 + \ldots A_n$ respectively.
Now, assume that $f = \phi_1 + \ldots + \phi_n$ and $g = \psi_1 + \ldots + \psi_m$.
Then, by functoriality, $[f] = [\phi_1] + \ldots + [\phi_n] = A_1 + \ldots + A_n$ and $g = [\psi_1] + \ldots + [\psi_m] = B_{1} + \ldots + B_{m}$. 
$f \not = g$ gives us two options.

@@

% \begin{theorem}

%     Let $\mathcal{R}$ be a rewriting system on $\textsf{PROP}(\Sigma)$ and $\langle l, r \rangle \in \mathcal{R}$.
%     $a \Rightarrow{}^{+}_{\langle l, r \rangle} b$ iff $[a] \Rrightarrow{}^{+}_{\langle [l], [r] \rangle} [b]$.
%     \begin{proof}
%         $a \Rrightarrow^{+}_{\langle l, r \rangle} b$ means that either
%         \begin{itemize}
%             \item $a$ and $b$ are decomposable as
%                   \[
%                     \scalebox{0.75}{
%                         \tikzfig{combinatorial_semantics/prop_rewriting}
%                     }
%                   \]
%                   \question{Here we can probably just reuse the proof going to $\ulcorner a \urcorner$ and then the result follows by functoriality of $[\cdot]$ (or $\llangle \cdot \rrangle$)}
%             \item or
%                  \[
%                     \scalebox{0.75}{
%                         \tikzfig{combinatorial_semantics/prop_plus_rewriting}
%                     }   
%                  \]
%                  and $a' \Rightarrow_{\langle l,r \rangle} b'$. This gives us the following decomposition in cospans
%                  % https://q.uiver.app/#q=WzAsMjcsWzAsNCwibSJdLFsxLDQsIkFfMSJdLFsyLDQsImkgKyBrIl0sWzMsNCwiOyJdLFs0LDAsImkiXSxbNSwwLCJYXzEiXSxbNiwwLCJqIl0sWzUsMiwiXFx2ZG90cyJdLFs1LDQsIkEnIl0sWzQsNCwiaSJdLFs2LDQsImoiXSxbNSwxLCIrIl0sWzUsMywiKyJdLFs1LDUsIisiXSxbNSw2LCJcXHZkb3RzIl0sWzUsNywiKyJdLFs0LDgsImkiXSxbNSw4LCJYX3ciXSxbNiw4LCJqIl0sWzcsNCwiXFxvdGltZXMiXSxbOCw0LCJrIl0sWzksNCwiayJdLFsxMCw0LCJrIl0sWzExLDQsIjsiXSxbMTIsNCwiaiArIGsiXSxbMTMsNCwiQV8yIl0sWzE0LDQsIm4iXSxbMCwxXSxbMiwxXSxbNCw1XSxbNiw1XSxbOSw4XSxbMTAsOF0sWzE4LDE3XSxbMTYsMTddLFsyMCwyMV0sWzIyLDIxXSxbMjYsMjVdLFsyNCwyNV1d
%                 \[
%                     \adjustbox{scale=0.5, center}{
%                 % \scalebox{0.75}{
%                     \begin{tikzcd}
%                     &&&& i & {X_1} & j \\
%                     &&&&& {+} \\
%                     &&&&& \vdots \\
%                     &&&&& {+} \\
%                     m & {A_1} & {i + k} & {;} & i & {A'} & j & \otimes & k & k & k & {;} & {j + k} & {A_2} & n \\
%                     &&&&& {+} \\
%                     &&&&& \vdots \\
%                     &&&&& {+} \\
%                     &&&& i & {X_w} & j
%                     \arrow[from=5-1, to=5-2]
%                     \arrow[from=5-3, to=5-2]
%                     \arrow[from=1-5, to=1-6]
%                     \arrow[from=1-7, to=1-6]
%                     \arrow[from=5-5, to=5-6]
%                     \arrow[from=5-7, to=5-6]
%                     \arrow[from=9-7, to=9-6]
%                     \arrow[from=9-5, to=9-6]
%                     \arrow[from=5-9, to=5-10]
%                     \arrow[from=5-11, to=5-10]
%                     \arrow[from=5-15, to=5-14]
%                     \arrow[from=5-13, to=5-14]
%                 \end{tikzcd}}
%                 \]
%                 and there is a convex match of $L$ in $A'$.
%                 It is then also down-closed and conflict reflecting as $A'$ is not hierarchical.
%                 Need to show that arising pushout complement is a boundary complement.
%                 \question{A problem is that the above terms does not decompose into a form of $A';C$ where $C$ is a pushout complement.
%                 This happens because we take composition along the external interfaces and this $A'$ is within a box, so it is not a part of an external interface.
%                 }
%                 \Aleksei{discuss this. We can simplify PROP+ rewriting to just a box, but still have a problem with context C}
%         \end{itemize}
%         \update{It is tempting to reuse a similar result for $\llbracket \cdot \rrbracket$ because according to our definition $f \Rrightarrow_{\mathcal{R}^{+}}$ means that there is an SMC piece $f'$ inside some box that rewrites to $g'$.
%             But we only have this result for $\llangle \cdot \rrangle$ for terms with bent wires which is a functor to $PROP + Frob$.
%         }
%     \end{proof}
% \end{theorem}

We will next consider functors
\[
    \textsf{PROP}(\Sigma, \mathcal{E}) \xrightarrow{\mathcal{J}} \textsf{PROP}^{+}(\Sigma, \mathcal{E}) \xrightarrow{\llangle - \rrangle} \sfrac{\WellTypedMdaEcospans}{\mathcal{S},\mathcal{E}}
\]
where $\mathcal{E}$ is a set of equations of the corresponding SMT.
\begin{theorem}[$\llangle - \rrangle$ is a functor]
\end{theorem}

\begin{theorem}[$\llangle - \rrangle$ is faithful]
\end{theorem}

\update{Or we can define rewriting for string diagrams in $\textsf{PROP}^{+}(\Sigma)$ based on rewriting in $\textsf{PROP}(\Sigma)$ and then say that $f \xrightarrow{}_{\mathcal{R}} g$ iff $[f] [g]$}

\begin{definition}[Rewriting in $\textsf{PROP}^{+}(\Sigma)$]

Recall the definition of rewriting in some $\textsf{PROP}(\Sigma) = \mathbb{A}$.
We say that $a \in \mathbb{A}$ rewrites into $b \in \mathbb{A}$ under a rewrite rule $\langle l, r \rangle$ where $l,r \in \mathbb{A}$ if $a$ and $b$ are decomposable as below.
\[
\scalebox{0.75}{
\tikzfig{combinatorial_semantics/prop_rewriting}
}
\]
and we will use notation $a \Rightarrow_{\langle l, r \rangle} b$.

Now consider an enriched prop $\mathbb{A}^{+}$ over the signature $\Sigma$ such that there is a functor $\mathcal{J} : \mathbb{A} \to \mathbb{A}^{+}$.
We than say that $a \in \mathbb{A}^{+}$ rewrites into $b \in \mathbb{A}^{+}$ under a rewrite rule $\langle l, r \rangle$ where $l,r \in \mathcal{J}(\mathbb{A})$ if $a$ and $b$ are decomposable as below.

\[
\scalebox{0.75}{
\tikzfig{combinatorial_semantics/prop_plus_rewriting}
}
\]

and $a' \Rightarrow_{\langle l, r \rangle} b'$.
We will use notation $a \Rightarrow_{\langle l, r \rangle}^{+} b$.
\end{definition}

\begin{theorem}
    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$.
\end{theorem}
